On Determinism of Game Engines used for Simulation-based Autonomous Vehicle Verification

### 2021-03-04 paper/cav_determinisim_main.tex
edits during review meeting

% In pursuit of an answer to this question we
In this paper we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
%
In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs. 
% KIE: This is covered in the next sentence:
% to identify conditions that result in non-determinism based on actor position variance in the simulation output.
%
\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified. }
%
% By analysing actor position variance 
We find that the CARLA simulator is non-deterministic under certain conditions. {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
%
CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected. {\color{red} Why ``vehicle'' collision and not just ``collision''?}
%
\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}

@@

% SEE BELOW \todo[inline]{Check the points on floating-point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
{\color{red} PROBLEMATIC: Even operations such as the average of an array can result in issues with overflows and underflows {\sc (these can happen with Integers too)}
and,  non-associative addition~\cite{Kapre2007} {\sc (only when the exec order is changed)} which may result in non-deterministic behaviour. }

@@

In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se. {\color{red} Should this be IMPRECISION rather than incorrectness?}
%
It is normally reasonable to assume that the processors on which game engines run are deterministic with respect to floating-point arithmetic. 
%
Thus, given the same operands and sequence of operations, the same processor should always produce the same bit pattern. %, i.e.\ a result that is as consistently equally \textit{incorrect} over a number of executions.
%
So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.

@@

\subsection{Actor Collisions}\label{S:Actor_Collisions}
Previous investigations into the Unreal Engine indicated that collisions between dynamic actors {\color{red} Why DYNAMIC?} and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation~\cite{collision_overview}, can lead to high simulation variance~\cite{TSLUnrealEngineTesting}. {\color{red} NOT ACCESSIBLE}
%
Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were identified as potentially key aspects to the investigation into simulation variance.

@@

% ***************************************************
%  Experiment Description
% ***************************************************
\subsection{Evaluation Metric}\label{s:Experiment_Description}
% \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
For each test the position of each actor is logged at $0.1s$ intervals providing a trace of that actor's path with respect to simulation time. 
%
The logs from repeated tests are compared at the same time index $t$ for each actor $a$ to provide a value for the variance of actor $a$ at time index $t$, $\sigma_a^{2}(t)$, at each point in time. {\color{red} This gives a variance function over time for each actor.}

{\color{blue} The logs from repeated tests are sourced to establish a value for the variance associated with each actor, $a$, at each time point $t$, giving a variance function over time for each actor, $\sigma_a^{2}(t)$.}

@@

The maximum variance over the entire set of $n$ repeated simulations, i.e.\ the overall observed worst case, is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}. 
%{\color{red} This is still a function over time - should it be?}
\begin{equation} \label{eq:max_sigma}
\max_{a,t}\sigma_a^{2}(t)
\end{equation}
The maximum deviation is the square root of the maximum variance {\color{red} The absolute value of the square root?} and herein referred to as ${\max\sigma}$ for brevity. {\color{red} So, ``a'' is any actor or a specific actor? Likewise, ``t'' is any time point or a specific time point? NOTE that any actor any time means a in A and t in T, where A is the set of actors and T is the set of all time points.}

@@

Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation. Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.\ \textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.\ \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}. Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating-point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking~\cite{releasing_project}.
% \todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings. The results were compared to a baseline of the default engine settings. 
%
However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values. Details on this previous investigation can be found on the Trustworthy Systems GitHub~\cite{TSLUnrealEngineTesting}. {\color{red} NOT ACCESSIBLE}

@@

A general pattern in the results indicates that some scenarios consistently fail to produce results within tolerance, irrespective of resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 2 \& 4 are above the dashed $1$cm line), while some are consistently within tolerance (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 5 \& 6 both are with pedestrians only), and some cases only fail to meet the tolerance requirement at higher resource utilisation levels, i.e.\ above 75\% resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 1 \& 3). {\color{red} ??? about 85\% based on the figure - or not???}

Examining specifically the results from Tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision. Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (Test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post collision. 
%
The pattern of permissible results prior to collision and non-permissible post collision is maintained up to a resource utilisation level of approximately 75\% {\color{red} Why 75\%?}, see Fig.~\ref{CarsCollisionPrePost}. This time series examination was repeated for vehicle to pedestrian collisions (Test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}. Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}. This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.

@@

% ***************************************************
%  METHOD
% ***************************************************
\section{Method to determine the Variance of a Simulation{\color{red} for what???}} \label{s:methodology}
In this section a method for assessing the actor path variance in AV simulation environments that are based on game engines is presented in the form of a work flow, see Fig.~\ref{method_diagram}. This method can be used to determine the operational domains of permissible variance of a simulation environment.  
%
{\color{red} MAKES NO SENSE: {\sc In this section a method for determining the variance of simulated actor path trajectories (KIE: a trajectory is a path), termed simulation variance, and resolving the operational domains of permissible variance of the simulation presented here as a work flow, see Fig.~\ref{method_diagram}.}}
%
In addition, recommendations and best practice guidelines for minimising simulation variance are suggested. {\color{red} \sc Highlight these in the text.}
% A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
% This methodology has been refined several times over the course of our investigations into both Unreal Engine and CARLA and is presented here as a work flow, see Fig.~\ref{method_diagram}. %These need to be considered at different stages in the process. 
%

@@

{\color{red} \sc Unclear how this relates to Analysis. Is this a summary of the method? If so, it should be moved to the front or to the conclusion.} This method can be extended to a broader actor states and actions including actor orientation, speed, and any other status indicators that may be of interest and also including appropriate actions that may be useful for verification purposes. This broader As such, Equation~\ref{eq:max_sigma} would need to be adjusted to include these new variables. {\color{red} \sc Something is missing in this sentence - please correct before moving to a different place, or removing.}

>>>>

% In pursuit of an answer to this question we
In this paper we investigate non-determinism and how it affects simulation results on the example of CARLA, an open-source autonomous driving simulation environment based on the Unreal game engine.
%
In our case study, scenarios between pedestrian and vehicle actors are investigated to determine the actor position variance in the simulation output for repeated simulation runs. 
% KIE: This is covered in the next sentence:
% to identify conditions that result in non-determinism based on actor position variance in the simulation output.
%
\todo[inline,color=red!40]{old: By limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a deterministic operational simulation setup could be identified. }
%
% By analysing actor position variance 
We find that the CARLA simulator is non-deterministic under certain conditions. {\color{red} Say what the observed variance range is, and what we consider permissible for this study.}
% However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.
%
CARLA exhibits a permissible variance when system utilisation is restricted to 75\% or less and the simulation is terminated once a vehicle collision has been detected. 
%
\todo[inline,color=green!40]{GC Alternative: By analysing actor position variance we find that the Carla simulator is non-deterministic under certain conditions. 
However, by limiting the level of system utilisation to less than 75\% and terminating scenarios once a collision has been detected, a permissible level of simulation variance was observed.}

@@

% SEE BELOW \todo[inline]{Check the points on floating-point operations in An Empirical Analysis of Flaky Tests~\cite{empirical-analysis-of-flaky-tests}.}
%{\color{red} PROBLEMATIC: Even operations such as the average of an array can result in issues with overflows and underflows {\sc (these can happen with Integers too)}
%and,  non-associative addition~\cite{Kapre2007} {\sc (only when the exec order is changed)} which may result in non-deterministic behaviour. }

@@

In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se. 
%{\color{red} Should this be IMPRECISION rather than incorrectness?}
%
It is normally reasonable to assume that the processors on which game engines run are deterministic with respect to floating-point arithmetic. 
%
Thus, given the same operands and sequence of operations, the same processor should always produce the same bit pattern. %, i.e.\ a result that is as consistently equally \textit{incorrect} over a number of executions.
%
So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.

@@

\subsection{Actor Collisions}\label{S:Actor_Collisions}
Previous investigations into the Unreal Engine indicated that collisions between actors and solid objects, termed \textit{blocking physics bodies} in Unreal Engine documentation~\cite{collision_overview}, can lead to high simulation variance~\cite{TSLUnrealEngineTesting}. 
%
Collisions and the subsequent physics calculations that are processed, termed \textit{event hit} callback in Unreal Engine, were identified as potentially key aspects to the investigation into simulation variance.

@@

% ***************************************************
%  Experiment Description
% ***************************************************
\subsection{Evaluation Metric}\label{s:Experiment_Description}
% \todo[inline]{Say what needs to be logged, what needs to be calculated based on these logs and what one should be looking for and why.}
For each test the position of each actor is logged at $0.1s$ intervals providing a trace of that actor's path with respect to simulation time. 
%
The logs from repeated tests are sourced to establish a value for the variance associated with each actor, $a$, at each time point $t$, giving a variance function over time for each actor, $\sigma_a^{2}(t)$.

@@

The maximum variance over the entire set of $n$ repeated simulations, i.e.\ the overall observed worst case, is defined as the largest variance of any actor at any time in any of the simulation runs, as given in Equation~\ref{eq:max_sigma}. 
%{\color{red} This is still a function over time - should it be?}
\begin{equation} \label{eq:max_sigma}
\max_{a,t}\sigma_a^{2}(t)
\end{equation}
The maximum deviation is the absolute value of the square root of the maximum variance and herein referred to as ${\max\sigma}$ for brevity. 
% {\color{red} So, ``a'' is any actor or a specific actor? Likewise, ``t'' is any time point or a specific time point? NOTE that any actor any time means a in A and t in T, where A is the set of actors and T is the set of all time points.}

@@

Within Unreal Engine there are numerous internal settings relating to the movement and interaction of physical bodies in the simulation. Settings can be adjusted to alter how actors interact and path plan via the navigation mesh of the environment, e.g.\ \textit{Contact Offset} and \textit{Navmesh Voxel Size}, or can be changed to improve the fidelity of physics calculations between game update steps, e.g.\ \textit{Physics Sub-Stepping} and \textit{Max Physics Delta Time}. Other options such as \textit{Enable Enhanced Determinism} were investigated along with running the engine from the command line with options for more deterministic behaviour \texttt{-deterministic}, floating-point control \texttt{/fp:strict} and headless mode \texttt{-nullrhi} along with running the test as a packaged release by building and cooking~\cite{releasing_project}.
% \todo[inline]{Maybe we can introduce our initial study here? Picture of the man and moving bloack and a result showing the path traces in x,y etc?}
An initial study into the Unreal Engine using a pedestrian and a moving block was used to investigate simulation variance against these settings. The results were compared to a baseline of the default engine settings. 
%
However, none of these options improved simulation variance significantly and all internal setting were set restored to the default values. Details on this previous investigation can be found on the Trustworthy Systems GitHub~\cite{TSLUnrealEngineTesting}.

@@

A general pattern in the results indicates that some scenarios consistently fail to produce results within tolerance, irrespective of resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 2 \& 4 are above the dashed $1$cm line), while some are consistently within tolerance (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 5 \& 6 both are with pedestrians only), and some cases only fail to meet the tolerance requirement at higher resource utilisation levels, i.e.\ above 75\% resource utilisation (cf.\ Fig.~\ref{ExperimentsStressSummary} Test 1 \& 3).

Examining specifically the results from Tests 2 \& 4 as a function of simulation time reveals further information about the simulation variance before and after an actor collision. Fig.~\ref{CarsCollisionCG25} shows this examination for vehicle to vehicle collisions (Test 2), where $\max\sigma$ switches from permissible prior to the vehicle collision to non-permissible post collision. 
%
The pattern of permissible results prior to collision and non-permissible post collision is maintained up to a resource utilisation level of approximately 75\%, see Fig.~\ref{CarsCollisionPrePost}. This time series examination was repeated for vehicle to pedestrian collisions (Test 4) and the results are shown in Fig.~\ref{CarsPeopleCollsionsCG25}. Similarly to vehicle-to-vehicle collisions, the variation of $\max\sigma$ for vehicle to pedestrian collisions indicates permissible pre-collision behaviour with up to 75\% resource utilisation, see Fig.~\ref{CarsPeopleCollisionPrePost}. This is a key finding; it suggests that verification engineers should consider terminating tests at the point of a collision, as any post-collision results will be non-permissible.

@@

% ***************************************************
%  METHOD
% ***************************************************
\section{Method to determine the Variance of a Simulation {\color{red} Say for what???}} \label{s:methodology}
In this section a method for assessing the actor path variance in AV simulation environments that are based on game engines is presented in the form of a work flow, see Fig.~\ref{method_diagram}. This method can be used to determine the operational domains of permissible variance of a simulation environment.  
%
{\color{red} MAKES NO SENSE: {\sc In this section a method for determining the variance of simulated actor path trajectories (KIE: a trajectory is a path), termed simulation variance, and resolving the operational domains of permissible variance of the simulation presented here as a work flow, see Fig.~\ref{method_diagram}.}}
%
In addition, recommendations and best practice guidelines for minimising simulation variance are suggested. {\color{red} \sc Highlight these in the text.}
% A number of parameters have been identified either theoretically or from the empirical investigation that influence simulation variance. 
% This methodology has been refined several times over the course of our investigations into both Unreal Engine and CARLA and is presented here as a work flow, see Fig.~\ref{method_diagram}. %These need to be considered at different stages in the process. 
%

@@
